---
title: "Project 1"
author: "Anna Brown, Rishika Deshmukh, and Ada Zhang"
date: "`r Sys.Date()`"
output:
     html_document:
          toc: true
          toc_depth: 2
          toc_float: true
          theme: cosmo
          code_folding: hide
---

# INTRODUCTION

This analysis delves into the relationship between a country's GDP and the issuance of F1 (student) and H1B (work) visas over time. By examining data across various countries and continents, we aim to uncover patterns that highlight the influence of economic performance on visa distribution. We hypothesize that higher GDP correlates with an increased number of F1 and H1B visas issued. Conversely, countries experiencing slower economic growth may see fewer visas granted.

# FOLLOW-UP QUESTIONS

Additionally, we can explore potential secondary questions, such as:

-   **How do presidential terms relate to the number of visa issuances?**

-   **What policies drive fluctuations in F1 and H1B visa approvals?**

# BACKGROUND

-   **Overview of U.S. Work Visas**

Types of Work Visa include H-1B(Specialty workers), L-1 (Intra-company transferees ), O-1(Individuals with extraordinary ability in their field), P-1,2,3, and etc. Our research focuses on H1B, the most populated U.S. work visa for highly skilled foreign professionals.

-   **What is H1B work visa?**

    The H-1B program allows U.S. employers to hire nonimmigrant workers for specialty occupations that require:

    -   Highly specialized knowledge.\

    -   At least a **bachelor’s degree** or its equivalent. (Dol.gov)

-   **H1B Policies change over time**

    -   1998 Industry Lobbying: High-tech industry representatives lobbied for an increase in the H-1B visa cap.

    -   In September 1997, the 65,000 annual cap for H-1B visas was reached for the first time. The cap provided new evidence of unmet demand for high-tech workers, particularly in the computer industry.

    -   In 1998, bills were introduced in the House and Senate to raise the H-1B cap from 65,000 to 115,000.

    -   Cap was lifted to 195,000 visas for 2001, 2002, and 2003 under the 2000 American Competitiveness in the 21st Century Act (AC-21).

    -   The annual H-1B cap reverted to 65,000 in 2004.

    -   To protect American workers during the recession, Congress targeted the H-1B program in the American Recovery and Reinvestment Act of 2009.

    -   The number of filed petitions fell by nearly 20 percent between 2007 and 2009.

    -   H-1B fees also fluctuate.

-   **Student Visa**

    -   F-1, J-1 (exchange programs), and M (nonacademic institution)

    -The F-1 visa allows a student to temporarily live in the United States for a defined period of time while studying at a school, college, seminary, or conservatory.

-   **Policies on F1 Student Visa**

    -   1996: Illegal Immigration Reform and Immigrant Responsibility Act (IIRIRA)

    -   Mandated the development of a program to collect information on nonimmigrant foreign students

    -   2002: Tightening of B Visa Regulations: regulations were amended to require individuals on B visas to change to F or M status before pursuing a course of study.

    -   2003: Implementation of SEVIS improving the tracking and monitoring of international students in the U.S

    -   Obama expanded and extended the federal Optional Practical Training program for STEM students to 29 months

    -   Under the Trump Administration(2017-2021), F-1 visa regulations became tighter.

        -   More rigorous evaluation of student visa applications and renewals.

        -   Proposed limitations on the length of stay, introducing fixed-term visa policies

-   **Unskilled Labor Visa**

-   H-2A, H-2B, H-2R, etc

-   Our main focus is H-2A and H-2B

    -H-2A: Agricultural workers (temporary labor for agricultural jobs that are seasonal or peak-period).

    -H-2B: Temporary non-agricultural workers (e.g., hospitality, construction, and landscaping work).

-   **Foreign Workers help U.S economy**

    -   Job Creation – H-1B workers fill skill gaps, boosting jobs and economic growth.

    -   Low Unemployment – Sectors with many H-1B workers show consistently low jobless rates.

    -   Innovation – H-1B workers drive patents, startups, and business expansion.

    -   High Wages – H-1B wages outpace U.S. averages, often exceeding prevailing rates. (americanimmigrationcouncil.org)

# METHODOLOGY

To analyze our initial problem, we used three data sources (originally accessed on January 2, 2025):

1.  [U.S. State Department](https://travel.state.gov/content/travel/en/legal/visa-law0/visa-statistics/nonimmigrant-visa-statistics.html){.btn .btn-primary}

Dataset Overview: The U.S. State Department's visa statistics provide detailed data on nonimmigrant visas issued by the U.S. government. It includes information about visa types and countries of origin. The dataset can be segmented by visa categories, geographic regions, and other criteria

Previous Uses: This dataset is widely used by researchers and policymakers to analyze trends in U.S. immigration, such as the number of visas issued to foreign nationals over time. The U.S. Department of Homeland Security and various research organizations use the visa statistics to evaluate the effects of visa policy changes. For instance, the Migration Policy Institute (MPI) uses this data to assess the impacts of changes in U.S. visa issuance on international student enrollment trends in U.S. universities.

2.  [World Bank](https://wits.worldbank.org/CountryProfile/en/country/by-country/startyear/ltst/endyear/ltst/indicator/NY-GDP-MKTP-CD#){.btn .btn-primary}

Dataset Overview: The World Bank dataset focuses on country-level economic indicators, specifically Gross Domestic Product (GDP) measured in current US dollars. The World Bank's World Integrated Trade Solution platform provides access to a wide array of economic, trade, and development data, which includes GDP data for all countries.

Previous Uses: The World Bank’s GDP data is widely used for forecasting global economic trends. For example, the World Economic Outlook regularly uses this data to produce global forecasts, guiding economic policies of member countries. Cross-country GDP comparisons are foundational in understanding global disparities in development. Dollar and Kraay (2003) used World Bank GDP data to argue that economic growth directly correlates with poverty reduction in developing nations.

3.  [U.S. Census Bureau](https://www.census.gov/data-tools/demo/idb/#/table?dashboard_page=country&COUNTRY_YR_ANIM=2025&menu=tableViz&quickReports=CUSTOM&CUSTOM_COLS=POP,GR,RNI,POP_DENS,CBR,CDR,NMR,NIM&TABLE_RANGE=1997,2014&TABLE_YEARS=1997,1998,1999,2000,2001,2002,2003,2004,2005,2006,2007,2008,2009,2010,2011,2012,2013,2014,2025&TABLE_USE_RANGE=Y&TABLE_USE_YEARS=Y&TABLE_STEP=1&TABLE_ADD_YEARS=2025){.btn .btn-primary}

Dataset Overview: The U.S. Census Bureau's International Data Base (IDB) provides demographic information for countries worldwide, including population size, growth rates, fertility rates, mortality rates, life expectancy, and migration patterns. The data spans multiple decades and is updated regularly. The IDB is valuable for understanding long-term trends in population and demographic characteristics across countries.

Previous Uses: Researchers have used the IDB to assess global population growth and predict demographic shifts. This is important for understanding the social and economic challenges faced by countries with rapidly growing populations, as well as the implications for urbanization, labor markets, and social services.

We used a Python Jupityer Notebook to clean each data frame, then merge them into a merged data frame for future analysis and vizualization using a common variable of country and year. All replication documents can be found within our public [GitHub repository](https://github.com/annab0503/DS4002/tree/main/Project%201/Analysis%20Data).

Variables in our (cleaned) dataset include:

| Variable                       | Description                                                                             | Source                |
|--------------------------------|-----------------------------------------------------------------------------------------|-----------------------|
| Country                        | geographic entity of interest                                                           | All                   |
| Fiscal Year                    | year for which data is reported, based on the U.S. fiscal calendar                      | All                   |
| Visa Category                  | classification of U.S. visas issued (e.g., unskilled labor, skilled labor, student)     | U.S. State Department |
| Quantity of U.S. Visas Granted | total number of visas issued for a given type and year                                  | U.S. State Department |
| Population                     | quantity of people living in a given geographic area                                    | U.S. Census Bureau    |
| GDP                            | Gross Domestic Product measured in current U.S. dollars                                 | U.S. Census Bureau    |
| Annual Growth Rate             | annual percent change in population, accounting for natural increase and net migration  | World Bank            |
| Rate of Natural Increase       | percent difference between the crude birth rate and the crude death rate                | U.S. Census Bureau    |
| Population Density             | total population of a geographic area divided by its land area in square kilometers     | U.S. Census Bureau    |
| Crude Birth Rate               | avg. annual number of births during a year per 1,000 population at midyear              | U.S. Census Bureau    |
| Net International Migrants     | number of immigrants minus the number of emigrants, including citizens and non-citizens | U.S. Census Bureau    |

## DATA CLEANING

#### LOADING PACKAGES

```{r message=FALSE}
library(tidyverse)
library(dplyr)
library(patchwork)
library(DT)
library(readr)
```

#### IMPORT DATA

```{r}
# Set the file path
url <- "https://raw.githubusercontent.com/annab0503/DS4002/main/Project%201/Analysis%20Data/analysis_data.csv"

# Read the CSV file from the URL
analysis_data <- read_csv(url)

# View the data
head(analysis_data)
```

#### ANALYSIS

```{r fig.width=12, fig.height=6, echo=FALSE, warning=FALSE}
# Define presidential administration transition years 
key_events <- c(2001, 2009, 2015, 2020)  # Adjust these as per your dataset

# Aggregate data by Fiscal Year and Visa Category, converting Total_Visas to units of 1000 
aggregated_data <- analysis_data %>% 
  group_by(`Fiscal Year`, `Visa Category`) %>% 
  summarise(Total_Visas = sum(`Quantity of U.S. Visas Granted`) / 1000, .groups = 'drop')  # Convert to thousands

# Custom color scheme for visa categories 
visa_colors <- c( 
  "Skilled Labor Visas" = "#F84947",  # Red 
  "Unskilled Labor Visas" = "#F1C40F",  # Golden Yellow 
  "Student and Exchange Visitor Visas" = "#5794D4"  # Blue 
)

# Create the stacked area chart with vertical indicators for presidential transitions 
ggplot(aggregated_data, aes(x = `Fiscal Year`, y = Total_Visas, fill = `Visa Category`)) + 
  geom_area() +  # Stacked area chart 
  geom_vline(xintercept = key_events, linetype = "dashed", color = "gray") + 
  scale_fill_manual(values = visa_colors) +  # Apply custom colors 
  labs(
    title = "Shifting Borders: The Impact of U.S. Political Events on Visa Issuance Trends", 
    x = "Fiscal Year", 
    y = "Total Quantity of Visas (in thousands)", 
    fill = "Visa Category" 
  ) + 
  theme_minimal() +  # Minimal theme for a clean look 
  theme(
    text = element_text(size = 12, family = "Futura"),  # Set font family to Futura 
    legend.position = "right",  # Position the legend on the right 
    plot.title = element_text(hjust = 0.5, family = "Futura", size = 16, margin = margin(b = 15)),  # Title customization 
    axis.title = element_text(family = "Futura"),  # Axis labels in Futura 
    axis.text = element_text(family = "Futura"),  # Axis text in Futura 
    legend.title = element_text(family = "Futura"),  # Legend title in Futura 
    legend.title.align = 0.5  # Center the legend title 
  ) + 
  # Add key event labels with adjustments for better visibility 
  annotate("text", x = key_events, y = Inf, label = c("9/11 Attack", "Great Recession", "Rep. Control of Congress", "COVID Pandemic"), 
           angle = 90, vjust = 1.5, hjust = 1, color = "black", size = 3, family = "Futura")
```

```{r fig.width=12, fig.height=6, echo=FALSE, warning=FALSE}
# Define presidential administration transition years
transition_years <- c(2001, 2009, 2017, 2021)  # Years of presidential transitions

# Aggregate data by Fiscal Year and Visa Category, converting Total_Visas to units of 1000
aggregated_data <- analysis_data %>%
  group_by(`Fiscal Year`, `Visa Category`) %>%
  summarise(Total_Visas = sum(`Quantity of U.S. Visas Granted`) / 1000, .groups = 'drop')  # Convert to thousands

# Custom color scheme for visa categories
visa_colors <- c(
  "Skilled Labor Visas" = "#F84947",  # Red for Skilled Labor Visas
  "Unskilled Labor Visas" = "#F1C40F",  # Golden Yellow for Unskilled Labor Visas
  "Student and Exchange Visitor Visas" = "#5794D4"  # Blue for Student and Exchange Visitor Visas
)

# Create the line chart with vertical indicators for presidential transitions
ggplot(aggregated_data, aes(x = `Fiscal Year`, y = Total_Visas, color = `Visa Category`, group = `Visa Category`)) + 
  geom_line(size = 1) + 
  geom_vline(xintercept = transition_years, linetype = "dashed", color = "gray") + 
  scale_color_manual(values = visa_colors) + 
  labs( 
    title = "From Bush to Biden: How Presidential Administrations Impact Types U.S. Visas Issued", 
    x = "Fiscal Year", 
    y = "Total Quantity of Visas (in thousands)", 
    color = "Visa Category" 
  ) + 
  theme_minimal() + 
  theme( 
    text = element_text(size = 12, family = "Futura"),  # Font family: Futura 
    legend.position = "right",  # Place the legend on the right 
    plot.title = element_text(hjust = 0.5, family = "Futura", size = 16, margin = margin(b = 15)),
    axis.title = element_text(family = "Futura"),  # Axis labels in Futura font 
    axis.text = element_text(family = "Futura"),  # Axis text in Futura font 
    legend.title = element_text(family = "Futura"),  # Legend title in Futura 
    legend.title.align = 0.5  # Center the legend title 
  ) + 
  annotate("text", x = transition_years, y = Inf, label = c("Bush", "Obama", "Trump", "Biden"), 
           angle = 90, vjust = 1.5, hjust = 1, color = "black", size = 4, family = "Futura")
```

# CONCLUSION

# WORKS CITED

-   <https://www.dol.gov/agencies/whd/immigration/h1b>

-   <https://ccis.ucsd.edu/_files/wp11.pdf>

-   <https://www.migrationpolicy.org/article/h-1b-temporary-skilled-worker-program>

-   <https://timelines.issarice.com/wiki/Timeline_of_student_visa_policy_in_the_United_States>

-   <https://www.thecrimson.com/article/2020/12/11/trump-immigration-higher-education/>

-   <https://www.americanimmigrationcouncil.org/research/h1b-visa-program-fact-sheet>

-   Migration Policy Institute. [Migration Policy Institute Website](https://www.migrationpolicy.org/)

-   Preston, S., & Elo, I. (1995). *Demography: Analysis of Mortality and Fertility Rates*. [DOI: 10.1016/S0304-3932(02)00206-4](https://doi.org/10.1016/S0304-3932(02)00206-4).
