---
title: "Project 1"
author: "Anna Brown, Rishika Deshmukh, and Ada Zhang"
date: "`r Sys.Date()`"
output:
     html_document:
          toc: true
          toc_depth: 2
          toc_float: true
          theme: cosmo
          code_folding: hide
---

# INTRODUCTION
This analysis delves into the relationship between a country's GDP and the issuance of F1 (student) and H1B (work) visas over time. By examining data across various countries and continents, we aim to uncover patterns that highlight the influence of economic performance on visa distribution. We hypothesize that higher GDP correlates with an increased number of F1 and H1B visas issued. Conversely, countries experiencing slower economic growth may see fewer visas granted. 

# METHODOLOGY
*Describe data cleaning/merging process and location of replication materials (link to github)*

To analyze our initial problem, we used three data sources (originally accessed on January 2, 2025):

1. [U.S. State Department](https://travel.state.gov/content/travel/en/legal/visa-law0/visa-statistics/nonimmigrant-visa-statistics.html){.btn .btn-primary}
- *review of dataset*
- *summary of background research/previous uses*

2. [World Bank](https://wits.worldbank.org/CountryProfile/en/country/by-country/startyear/ltst/endyear/ltst/indicator/NY-GDP-MKTP-CD#){.btn .btn-primary}
- *review of dataset*
- *summary of background research/previous uses*

3. [U.S. Census Bureau](https://www.census.gov/data-tools/demo/idb/#/table?dashboard_page=country&COUNTRY_YR_ANIM=2025&menu=tableViz&quickReports=CUSTOM&CUSTOM_COLS=POP,GR,RNI,POP_DENS,CBR,CDR,NMR,NIM&TABLE_RANGE=1997,2014&TABLE_YEARS=1997,1998,1999,2000,2001,2002,2003,2004,2005,2006,2007,2008,2009,2010,2011,2012,2013,2014,2025&TABLE_USE_RANGE=Y&TABLE_USE_YEARS=Y&TABLE_STEP=1&TABLE_ADD_YEARS=2025){.btn .btn-primary}
- *review of dataset*
- *summary of background research/previous uses*

We used a python jupityer notebook to clean each data frame, then merge them into a merged data frame for future analysis and vizualization. All replication documents can be found within our public [GitHub repository](https://github.com/annab0503/DS4002/tree/main/Project%201/Analysis%20Data).

Variables in our (cleaned) dataset include:

| Variable                       | Description                                       |
|--------------------------------|---------------------------------------------------|
| Country                        | Description                                       |
| Fiscal Year                    | Description                                       |
| Type of U.S. Visa              | Description                                       |
| Quantity of U.S. Visas Granted | Description                                       |
| Population                     | Description                                       |
| GDP (Current USD)              | Description                                       |
| Annual Growth Rate             | Description                                       |
| Rate of Natural Increase       | Description                                       |
| Population Density             | Description                                       |
| Crude Birth Rate               | Description                                       |
| Net International Migrants     | Description                                       |

## DATA CLEANING
#### LOADING PACKAGES
```{r message=FALSE}
library(tidyverse)
library(dplyr)
library(patchwork)
library(DT)
```

#### IMPORT DATA
```{r}
# Set the file path
file_path <- "/Users/annabrown/Downloads/merged_data.csv"

# Import data frame
df <- read.csv(file_path)

# View imported data frame
head(df)
```
#### GENERAL CLEANING

#### Analysis
# CONCLUSION
